{\LARGE \textsc{Structural controlability of pollination networks}}

### 2016-05-06 paper/manuscript.Rnw
revised methods and results

The number of nodes necessary to fully control a complex network can be calculated by counting the number of unmatched nodes in the network's maximum matching \autocite{Liu2011}. In a directed network a matching is a subset of links in which no two links share a common starting node or a common ending node; a node is matched if it is the ending node of one of the links in the matching (\autoref{fig:explain_mm}a). A matching is then maximum if \textit{(i)} the number of matched liks---also known as matching size---is maximal and \textit{(ii)} the sum of the weights of the matched links---known as matching weight---is the largest possible among all possible matchings \autocite{West2001}. To find the maximum matching we used an alternative bipartite representation of the network in which the two levels indicate the outgoing and incoming links to each node (\autoref{fig:explain_mm}b). Finding a matching in this alternative representation is equivalent to finding the set of links such that each node on the top level is matched to at most one node on the bottom level, and vice versa \autocite{West2001, Csardi2006a}.

@@

multiplot(p1, p2, layout = matrix(nrow = 1, c(1,1,1,2,2,2,2,2)))
@
\caption{\label{fig:explain_mm}  One of the possible maximum matchings of a network (dark grey links) defines a set matched ($p_1$, $p_2$, and $a_3$) and unmatched nodes ($a_1$ and $a_2$). The minimum number of species neccessary to control this network equates to the number of unmatched nodes $n_d = 2$. Superior nodes (dark grey circles) drive the dynamics of other species and are therefore important for the controlability of the network because they either require external input or because the are part of the chain that transmit the external imputs to other species.}
\end{figure}

Given a maximum matching, the link to network controllability has a relatively straightforward and intuitive basis. In control theory, a network is fully controlable when each node has a "superior", this is, an input that drives its dynamics. Matched nodes can thus be controled by the node pointing to it---its superior. Unmatched nodes on the other hand, have no superior in the network and must be directly controlled by external inputs \autocite{Liu2011}. In an ecological context, this is what we refer to as management interventions.

Systems that require a large number of external inputs are more difficult or costly to control. The \textit{manageability} of a community $m$ is therefore determined by the proportion $n_d$ of species on which we would need to apply management interventions in order to gain full control of the abundances of all species in the ecological community, $$m = 1- \frac{n_d}{s},$$ where $s$ is the total number of species in the community.

\subsection*{Weighting links}

Recall that our maximum matching algorithm requires a directed network in which a link between species $i$ and $j$ pointing to species $j$ indicates that the abundance of $j$ can be affected by the abundance of $i$. In ecological networks, establishing a directionality for the links between species is relatively straightforward when they represent resources or energy flow, however, interspecific effects---particularly in pollination networks---are not strictly directed. Nevertheless, the extent to which a pair of interacting species affect each other can be quantified by the magnitude of the mutual dependence \autocite{Bascompte2006}. As the dependencies are bidirectional, this assumption renders a bipartite network in which interacting pairs are connected by two directed links (\autoref{fig:explain_cd}b).

@@

weights <- allEdges %>%
  dplyr::filter(Sequence == 12) %>%
  dplyr::mutate(weight = c(0.75, 1, 0.5, 0.6, 1, 0.16))

@@

@
\caption{\label{fig:explain_cd} Mutualistic networks are usually described quantitatively by the relative number of visits between each plant and animal species (a). The mutual dependencies between interacting species (b) depend directly on the relative visitation frequency. The dependency of plant $i$ on pollinator $j$, $d_{ij}$, is the proportion of the visits coming from pollinator $j$ compared to all pollinator visits to plant $i$. Likewise, the dependency of pollinator $j$ on plant $i$, $d_{ji}$, is the ratio of the visits by pollinator $j$ to plant $i$ and all visits of pollinator $j$. Nevertheless, mutual dependencies in pollination networks tend to be highly assymetric. The relative differences of dependencies (c) then is a good indicator  }
\end{figure}

Nevertheless, mutual dependencies have been shown to be highly assymetric in natural communities \autocite{Bascompte2006}. In other words, if a plant species is largely dependent on a pollinator species, then the pollinator tends to depend only very weakly on the plant. This suggest that it may be possible to relax our initial assumption and simplify the network such that interacting species are connected by one directed link when mutual dependencies are highly assymetric (\autoref{fig:explain_cd}c). The extent to which a species $i$ affects species $j$ relative to the extent to which species $j$ affects species $i$ can then be summarised by the interaction asymmetry.

$$a(i,j) = \frac{\left | d_{ij}-d_{ji} \right |}{\textup{max}\left ( d_{ij}, d_{ji} \right )}$$

@@

\todo[inline]{insert ecological motivation?}

The manageability of a community $m$ provides an indication of the magnitude of $n_d$ but not the identity of the species that compose this set. Moreover, maximum matchings are often not unique in for a given network and different sets of species of size $n_d$ could be used to control the network \autocite{Liu2011}. Species vary in terms of the frequency with which they are classified as superior nodes (\autoref{fig:explain_mm}), and this frequency provides an indication of their relative importance in driving the state of the community.

We therefore call \textit{driver species} those that are identified as being a superior node in all\todo{check}  possible matchings, and consequently likely to have a disproportionate effect on the abbundances of other species in the community. Unlike unweighted networks, where a maximum matching is defined by its matching size (the number of matching links), a maximum matching in a weighted network is one whose links have the largest possible total weight among all possible matchings. This means that in weighted networks there is a number of matchings that have maximal size, but are not maximum matchings because their total weight is not maximal.

Consequently, we identified the driver species in each of the networks by computing all possible matchings in the network that have the same maximal size while allowing for some tolerance regarding the weight. We relax the matching weight requirement because empirical intreaction strengths, from which we calculate dependencies, are stochastic to some extent and depend on the sampling method and intensity. To compute all the matchings with maximal size, we first generated the network's line graph---an alternative representation in which nodes are the interactions of the original network and links represent common ending points. We then calculated the complement of the line graph---a graph in which nodes are adjacent if and only if they are not adjacent in the line graph. All possible maximum matchings can be found by enumerating all cliques of size $1-n_d$ in this complement graph \autocite{West2001, Csardi2006a}.

@@

In the last section we outlined the theoretical and computational underpinings of our approach. Now we illustrate its empirical application using ten pairs of plant-pollinator communities for which we calculated their manageability and the relative importance of species. Each network pair was composed by a community invaded by a plant and a community effectively free of the invasive species (Table S1). The networks were constructed from previously published visitation data collected from pollination communities in Bristol, UK \autocite{Lopezaraiza-Mikel2007} and in Cap de Creus National Park, Spain \autocite{Bartomeus2008}. In the four British uninvaded communities, the non-invaded plots were obtained by experimentally removing all the flowers of the invasive species \textit{Impatients grandulifera}. In contrast, the Spanish uninvaded communities were obtained from plots that had not yet been colonised by either of the invasive species \textit{Carpobrotus affine acinaciformis} or \textit{Opuntia stricta}. The structure of these networks was determined using visitation frequency which has been shown to be an appropriate surrogate for inter-specific effects in pollination networks \autocite{Vazquez2005, Bascompte2006}. To examine the importance of this decision, we also evaluated the effect of using pollinator efficiency or importance as alternative measures of species interactions \autocite{NeEman2010, Ballantyne2015}, and we found consistent results for all of these options(see Supplementary Information).

@@

We started by quantifying the manageability of the analysed pollination communities. To do that, we calculated the maximum matching of the corresponding pollination network, and estimated the minimum number of species that need directed intervecntions to fully control the species abbundances in the community. If a network had more than one component (two species are in different components if there exist no path between them) we analysed the largest one. Although understanding the distribution of manageability across ecological networks is an useful result in itself, we wanted to test wheter perturbations, like invasion status, have an impact on manageability values. As, naively one might expect to see a negative relationship between diversity and manageability \autocite{Menge1995}, we used a set of generalised linear models (with binomial error structure) to investigate the effect of invasion status while controling for \textit{(i)} the total number of species, \textit{(ii)} plant richness, \textit{(iii)} polinators richness, \textit{(iv)} the ratio of plants to pollinators richness, and \textit{(v)} the study site. Competing models were assessed by comparing the Akaike Information Criterion corrected for small sample sizes (AICc).

Subsequently we explored the extent to which structural factors in the empirical networks influence their manageability. Previous studies found a direct link between a network's degree distribution and the number of nodes necessary to fully control it \autocite{Liu2011}, but this relationship has not been explored in weighted ecological networks. We therefore compared the manageability of the empirical networks to networks generated by three null models. One set of random visitation networks was generated by maintaining each species' degree (its number of interactions) while varying the number of visits per species. A second set of random networks mantained the species' strength (its total sum of visits) while varying their degrees. Randomisations were generated using vegan 2.3-3 \autocite{Oksansen2016}. Finally a three set of fully random networks maintained only the connectivity of the empirical network, but both degree and weight were allowed to vary. We then calculated the mutual dependences and interaction asymmetries of the random networks and calculated their manageability using the maximum matching algorithm. Finally we calculated the p-value of the manageability of each empirical network and the corresponding manageability of 999 randomisations.

@@

Our third question is related to how different species differ on their importance for the population dynamics of the community. To quantify this importance we computed all possible matchings that have the same size as the maximum matchings and in each network. We then calculated the frequency to which each species is deemed to be a superior node in matchings with a weight equal or larger to 0.5 times the weight of the maximum matchings. We choose this threshold as it provided a high agreement between the relative importance of species calculated by weighting interactions with the mutial dependencies or the depndence asymmetry as well as between the network is quantified by visitation or pollination efficiency. Nevertheless, the choice of the threshold only had a negligible impact on the results (Supplementary information).

The number of maximal matchings increasses exponentially with the number of links and nodes in a network. When we use mutual dependecies, the large number of links in the most diverse communities renders the computation unfeasible. Taking this limitation into account we compared the relative network manageability obtained when using assimetries or when using mutual dependencies for eleven networks---for which the number of matchings was smaller than $4\times 10^{10}$. Results from a paired Spearman correlation test, the weighted correlation coefficient $r_{w2}$---which gives more weight to higer ranks \autocite{PintoDaCosta2015}, and the Jaccard similarity of the driver species in both approaches, suggested a high similarity in relative ranks. Therefore, we used the results from the assymetries, rather than mutiual dependences, to further investigate the factors that determine species' relative importance.

We then asked the question of whether some species-level structural properties can predict the relative importance of species. Here, we used a Generalised Linear Mixed-Effects model to evaluate the effect on the species' relative importance of a suite of structural measures. We included measures of centrality (degree and eigen-centrality) which have been found to be strong predictors of importance in a coextintion context\todo{add ref}; measures related to network robustness (contribution to nestedness) as nestedness has been proposed as one of the key properties that promote stability in mutualistic networks \autocite{Saavedra2011}; and measures of strength of association and depedency (visitation and dependency strength) as their skewed distribution is characteristic of plant-pollinator networks and determine the extent of interspecific effects. All metrics were calculated using bipartite 2.06 \autocite{Dormann2008}. To allow for variation among different ecological communities, this factor was included as a random effect \autocite{Bates2015a}. Candidate models were compared using AIC and the relative importance of the explanatory variables was evaluated using the sum of Akaike weights over candidate models that accounted for 95\% of the evidence \autocite{Burnham2003, Bates2014, Barton2016}.

Invasive species have been shown to have disproportionate effects on the communities they invade. So we examined the extent to which invasive species tend to be driver species---species whith the maximum importance. As our invasive species were plants, we compared the relative importance of invasive species to those of other plants in the community and to which degree they differ when accounting for the structural factors that determine their importance\todo{needs to be developed once the results section is rewritten}.

>>>>

The number of nodes necessary to fully control a complex network can be calculated by counting the number of unmatched nodes in the network's maximum matching \autocite{Liu2011}. In a directed network, a given matching is a subset of links in which no two links share a common starting node or a common ending node; a node is matched if it is the ending node of one of the links in the matching (\autoref{fig:explain_mm}a). A matching is then maximum if bith \textit{(i)} the number of matched links---known as matching size---is maximal and \textit{(ii)} the sum of the weights of the matched links---known as matching weight---is the largest possible among all possible matchings of that size \autocite{West2001}. To find the maximum matching, we used an alternative bipartite representation of the network in which the two levels indicate the outgoing and incoming links to each node (\autoref{fig:explain_mm}b). Finding a matching in this alternative representation is equivalent to finding a set of links such that each node on the top level is matched to at most one node on the bottom level, and vice versa \autocite{West2001, Csardi2006a}.

@@

multiplot(p1, p2, layout = matrix(nrow = 1, c(1,1,1,2,2,2,2,2)))
@
\caption{\label{fig:explain_mm} Finding a maximum matching in a complex network. a) We start with a directed network  in which the direction of the link represent the direction of control. This is, a link from $a_1$ to $p_1$ indicates that the state of $p_1$ (density/abbundances in an ecological context) are influenced by the state of $a_1$. b) To facilitate the computation of the maximum matching, we used an alternative undirected bipartite representation of the directed network in which each node in (a) is represented by two nodes that indicate the outgoing and incoming links. One of the possible maximum matchings of a network (dark grey links) defines a set matched ($p_1$, $p_2$, and $a_3$) and unmatched nodes ($a_1$ and $a_2$). The minimum number of species neccessary to control this network equates to the number of unmatched nodes $n_d = 2$. Superior nodes (dark grey circles) drive the dynamics of other species and are therefore important for the controlability of the network because they either require external input or because the are part of the chain that transmit the external imputs to other species.}
\end{figure}

Given a maximum matching, the link to network controllability has a relatively straightforward and intuitive basis. In control theory, a network is fully controlable when each node has a ``superior'', that is, an input that drives its dynamics. Matched nodes can thus be controled by the node pointing to it. Unmatched nodes, on the other hand, have no superior in the network and must be directly controlled by external inputs \autocite{Liu2011}. In an ecological context, this is what we refer to as management interventions.

Systems that require a large number of external inputs are more difficult or costly to control. Therefore, the manageability $m$ of a community can be approximated by the proportion of species on which we would need to apply management interventions in order to gain full control of the abundances of \textit{all} species in the ecological community, $$m = 1- \frac{n_d}{s},$$ where $n_d$ is the proportion of species needed to gain full control and $s$ is the total number of species in the community.

\subsection*{Weighting \& directing links}

Recall that our maximum matching algorithm requires a directed network in which a link between species $i$ and $j$ pointing to species $j$ indicates that the abundance of $j$ can be affected by the abundance of $i$. This implies that we need first to establish a directionality for the links between species that in consistent with the dynamics of the comunity. In some ecological networks, establishing a directionality is relatively straightforward when they represent resources or energy flow. Interspecific effects in pollination networks, however, are not strictly directed. Nevertheless, the extent to which a pair of interacting species affect each other can be quantified by the magnitude of the mutual dependence \autocite{Bascompte2006}.

The dependency of plant $i$ on pollinator $j$, $d_{ij}$, is the proportion of the visits coming from pollinator $j$ compared to all pollinator visits to plant $i$. Likewise, the dependency of pollinator $j$ on plant $i$, $d_{ji}$, is the ratio of the visits by pollinator $j$ to plant $i$ and all visits of pollinator $j$. As the dependencies are bidirectional, adopting this assumption generates a weighted bipartite network in which interacting pairs are connected by two directed links (\autoref{fig:explain_cd}b).

Mutual dependencies have been shown to be highly assymetric in natural communities \autocite{Bascompte2006}. In other words, if a plant species is largely dependent on a pollinator species, then the pollinator tends to depend only very weakly on the plant. This suggest that it may be possible to relax this initial assumption and simplify the network further such that interacting species are only connected by one directed link when mutual dependencies are assymetric (\autoref{fig:explain_cd}c). The extent to which a species $i$ affects species $j$ relative to the extent to which species $j$ affects species $i$ can then be summarised by the interaction asymmetry $$a(i,j) = \frac{\left | d_{ij}-d_{ji} \right |}{\textup{max}\left ( d_{ij}, d_{ji} \right )}$$

Employing an unidrectional interaction asymmetry, as a surrogate of bidirectional mutual dependences, dramatically reduces the number of links in the network. Which damatically simplifies the computation of of all posible maximum matchings in a network.

@@

weights <- allEdges %>%
  dplyr::filter(Sequence == 12) %>%
  dplyr::mutate(weight = c(0.75, 0.5, 0.5, 0.6, 0.5, 0.16))

@@

@
\caption{\label{fig:explain_cd} Different ways to depict quantitative mutualistic networks. a) Pollination networks are usually described by the number of visits between each plant and animal species. b) Given visitation data the mutual dependencies between interacting species depend directly on the relative visitation frequency. c) The relative differences of dependencies---the interaction asymmetry---then provides a mean to simplify the interspecific effects.}
\end{figure}

@@

The manageability of a community $m$ provides an indication of the magnitude of $n_d$ but not the identity of the species that compose this set. Moreover, maximum matchings are often not unique in for a given network and different sets of species of size $n_d$ could be used to control the network \autocite{Liu2011}. Species vary in terms of the frequency with which they are classified as superior nodes (\autoref{fig:explain_mm}), and this frequency provides an indication of their relative importance in driving the state of the community. Ecologically, these distinctions are crucial because resources are limited and therefore ecological inteventions should be focused on species that might provide the largest impact. We therefore call \textit{driver species} those that are identified as being a superior node in all possible matchings, and consequently likely to have the most disproportionate effect on the abundances of other species in their community.

As previously described, unlike unweighted networks, where a maximum matching is defined solely by its matching size, a maximum matching in a weighted network is one whose links have the largest possible total weight among all possible matchings. This means that in weighted networks there is a number of matchings that have maximal size, but are not maximum matchings because their total weight is not maximal. One alternative to calculate the species importance, and identify the driver species, would be to consider only maximum matchings. However empirical intreaction strengths, from which we calculate dependencies, are to some extent stochastic and depend on the sampling method and intensity. We therefore relaxed the matching weight requirement computed all possible matchings in the network that have the same maximal size while allowing for some tolerance regarding the weight. To compute all the matchings with maximal size, we first generated the network's line graph---an alternative representation in which nodes are the links of the original network and links represent common ending points\todo{add ref}. We then calculated the complement of the line graph: a graph in which nodes are connected if and only if they are not connected in the line graph. All possible matchings of maximal size can be found by enumerating all cliques of size $1-n_d$ in this complement graph \autocite{West2001, Csardi2006a}.

@@

In the last section we outlined the theoretical and computational underpinings of our approach. Now we illustrate its empirical application using ten pairs of plant-pollinator communities for which we explored their manageability and the relative importance of their constituent species. The networks were constructed from previously published visitation data collected from pollination communities in Bristol, UK \autocite{Lopezaraiza-Mikel2007} and in Cap de Creus National Park, Spain \autocite{Bartomeus2008}. Notably, each network pair was composed of a community invaded by a plant and a community ``free'' of the invasive species (Table S1). In the four British uninvaded communities, the non-invaded plots were obtained by experimentally removing all the flowers of the invasive species \textit{Impatients grandulifera}. In contrast, the Spanish uninvaded communities were obtained from plots that had not yet been colonised by either of the invasive species \textit{Carpobrotus affine acinaciformis} or \textit{Opuntia stricta}. The structure of these networks was determined using visitation frequency which has been shown to be an appropriate surrogate for interspecific effects in pollination networks \autocite{Vazquez2005, Bascompte2006}. To examine whether this decision would influence our results, we also evaluated the effect of using pollinator efficiency or importance as alternative measures of species interactions \autocite{NeEman2010, Ballantyne2015}, and we found quantitatively results for all of these options(see Supplementary Information).

@@

Given the different networks, we started by quantifying their manageability. To do that, we calculated the maximum matching of the corresponding pollination network, and estimated the minimum number of species that need directed intervecntions to fully control the species abbundances in the community. If a network had more than one component (two species are in different components if there exists no path between them), we only analysed the largest one. Although understanding the variability of manageability across ecological networks is an useful result itself, we also wanted to test whether perturbations, like invasion status, have an impact on manageability values. We therefore used a set of generalised linear models (with binomial error structure) to investigate the effect of invasion status while controling for factors related to species richness---as naively one might expect to see a negative relationship between diversity and manageability \autocite{Menge1995}. Coovariates inlcuded the total number of species, plant richness, polinators richness, the ratio of plants to pollinators richness, and the study site. We assessed competing models by comparing the Akaike Information Criterion corrected for small sample sizes (AICc).

We next explored the extent to which structural properties of the empirical networks influence their manageability. Previous research indicates a direct link between a network's degree distribution and the number of nodes necessary to fully control it \autocite{Liu2011}, but the strength and applicability of this relationship has not been tested for in weighted ecological networks. We therefore compared the manageability of the empirical networks to networks generated by three null models which varied in terms of the constraints they impose. The first set of random visitation networks maintain the connectivity of the empirical network, but both degree and weight were allowed to vary. A second set set was generated by maintaining each species' degree (its number of interactions) while varying the number of visits per species. Finally a third set of random networks mantained the species' strength (its total sum of visits) while varying their degrees. Randomisations were generated using vegan 2.3-3 \autocite{Oksansen2016}. After generating the randomized networks under these three schemes, we then calculated the mutual dependences and interaction asymmetries and determined their manageability using the maximum matching algorithm. Finally we calculated the p-value of the manageability of each empirical network and the corresponding manageability of 999 randomisations.

@@

Our second key question is related to how different species differ in their importance for the population dynamics of the community. To quantify this importance in each network, we computed all possible matchings that have the same size as the networks' maximum matching. We then calculated the frequency to which each species is deemed to be a superior node in matchings with a weight greater or equal to 0.5 times the weight of the maximum matching. We choosed this threshold as it provided a high agreement between networks quantified by visitation and pollination efficiency as well as between our weighting/directionality assumptions (mutual dependences and inetraction asymmetries). Note that the choice of this threshold had a negligible impact on the results (Supplementary information).

We observed that the number of maximal matchings increasses exponentially with the number of links and nodes in a network. Unfortunately, this implies that the large number of links in the most diverse communities renders the computation unfeasible when we use mutual dependecies. Taking this limitation into account, we compared the relative network manageability obtained when using assimetries or when using mutual dependencies for eleven of the twenty networks---for which the number of matchings was smaller than $4\times 10^{10}$. The comparison was performed using paired Spearman correlation test, a weighted correlation coefficient $r_{w2}$---which gives more weight to higer ranks \autocite{PintoDaCosta2015}, and a Jaccard similarity of the driver species in both approaches. All three approaches suggested a high similarity in relative ranks. Therefore, we used the results from the assymetries, rather than mutual dependences, to further investigate the factors that determine species' relative importance.

Given the species frequency obtained using the interaction asymmetries, we then asked the question of whether some species-level structural properties can predict the relative importance of species. To do so, we used a generalised linear mixed-effects model (GLMM) to evaluate the relationship between the species' relative importance and a suite of structural measures. We specifically included measures of centrality (degree and eigen-centrality) which have been found to be strong predictors of importance in a coextintion context\todo{add ref}; measures related to network robustness (contribution to nestedness) as nestedness has been proposed as one of the key properties that promote stability in mutualistic networks \autocite{Saavedra2011}; and measures of strength of association and depedence (visitation and dependency strength), as their skewed distribution is characteristic of plant-pollinator networks and determine the extent of interspecific effects. All metrics were calculated using the R package bipartite 2.06 \autocite{Dormann2008}. We allowed for variation among different communities by including the network identity as a random effect \autocite{Bates2015a}. Candidate models were compared using AICc and the relative importance of the explanatory variables was evaluated using the sum of Akaike weights over candidate models that accounted for 95\% of the evidence \autocite{Burnham2003, Bates2014, Barton2016}.

Invasive species have been shown to have disproportionate effects on the communities they invade\todo{add ref}. So we lastly examined the extent to which invasive species tend to be driver species---species whith the maximum importance---and to which extent their importance could be explained by the factors previously described. To do so, as our invasive species were plants, we first used a Mann-Withney U test to compare the relative importance of invasive species to those of other plants in the community. We then examined the residuals of the GLMM to investigate if invasive species' importance is larger or smaller than expected given their network structural properies.
